---
title: "population data"
subtitle: "BC Data Catalogue"
author: "Martin Monkman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# packages
library(tidyverse)
library(here)

library(bcdata)

```

## read data

Source:

-   https://www2.gov.bc.ca/gov/content/data/statistics/people-population-community/population/population-estimates

-   https://www2.gov.bc.ca/assets/gov/data/statistics/people-population-community/population/pop_municipial_subprov_areas.xlsx

Source: BC Stats, Population Estimates by Regional Districts and Development Regions

Population Distribution Vancouver Island and Coast by Age Group\
2011 2016 2021\
Under 15 years old 14% 13% 13% Working Age Population (Aged 15 - 64) 67% 65% 62% Population 65+ 19% 22% 25%

Source: BC Stats, Population by Age and Sex.

#### download

https://bcstats.shinyapps.io/popApp/

Select:

-   region type == "Regional District"

-   region(s) == as above

-   years == all

-   genders == Totals

-   age format == "5-year Age Groups"

-   variables to display as columns = Age

Rename file as "pop_regional district_est proj_5 yr.csv"


NOTE: StatCan table with same data for Canada (but no projections) is 

Census Division

Population estimates, July 1, by census division, 2016 boundaries1, 2, 3, 4, 5, 6
Frequency: Annual

Table: 17-10-0139-01

Release date: 2023-01-11

Geography: Canada, Province or territory, Census division
https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710013901

Census Subdivision: https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710014201


#### direct pull!

Uses BC Data Catalogue

```{r}
bcdata::bcdc_get_record("86839277-986a-4a29-9f70-fa9b1166f6cb")

# list all available population files by geography
#pop_data <- bcdata::bcdc_get_data("86839277-986a-4a29-9f70-fa9b1166f6cb")

# Economic Region (Development Region)
data1 <- bcdc_get_data('86839277-986a-4a29-9f70-fa9b1166f6cb', resource = '6ec0805b-2f6f-4b63-95e8-d3a818469622')
file_name <- glue::glue("pop_bcdata_devreg_{today()}.rds")
write_rds(data1, here::here("data_output", file_name))

# regional district (Census division)
data1 <- bcdc_get_data('86839277-986a-4a29-9f70-fa9b1166f6cb', resource = '36610a52-6f90-4ed6-946d-587641a490df')
file_name <- glue::glue("pop_bcdata_rd_{today()}.rds")
write_rds(data1, here::here("data_output", file_name))


```

```{r data_pull, eval=FALSE}

resource_codes <-  bcdata::bcdc_tidy_resources("86839277-986a-4a29-9f70-fa9b1166f6cb") %>% pull(id)

data1 <- map_dfr(
resource_codes,
bcdata::bcdc_get_data,
record = "86839277-986a-4a29-9f70-fa9b1166f6cb",
show_col_types = FALSE
)

file_name <- glue::glue("pop_bcdata_{today()}.rds")
write_rds(data1, here::here("data_output", file_name))

```

```{r data_read}
file_name <- glue::glue("pop_bcdata_{today()}.rds")

data1 <- read_rds(here::here("data_output", file_name))

```

explore the data 

```{r}
unique(data1$Region.Type)
```


```{r}
df_rd_pop_total <- data1 |> 
  filter(Region.Type == "Regional District") |> 
  filter(Gender == "T") |> 
  select(Region:Total)

df_rd_pop_total
```
