---
title: "population map & pyramid"
subtitle: "Comox Valley RD"
author: "Martin Monkman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# packages
library(tidyverse)
library(here)

```

## read data

Source:

-   https://www2.gov.bc.ca/gov/content/data/statistics/people-population-community/population/population-estimates

-   https://www2.gov.bc.ca/assets/gov/data/statistics/people-population-community/population/pop_municipial_subprov_areas.xlsx

Source: BC Stats, Population Estimates by Regional Districts and Development Regions

Population Distribution Vancouver Island and Coast by Age Group\
2011 2016 2021\
Under 15 years old 14% 13% 13% Working Age Population (Aged 15 - 64) 67% 65% 62% Population 65+ 19% 22% 25%

Source: BC Stats, Population by Age and Sex.

#### download

https://bcstats.shinyapps.io/popApp/

Select:

-   region type == "Regional District"

-   region(s) == as required

-   years == all

-   genders == M & F

-   age format == "Single Year Age Groups"

-   variables to display as columns = Age

Rename file as "Comox_Valley_Population_Projections_2024-02-04.csv"


NOTE: StatCan table with same data for Canada (but no projections) is 

Census Division

Population estimates, July 1, by census division, 2016 boundaries1, 2, 3, 4, 5, 6
Frequency: Annual

Table: 17-10-0139-01

Release date: 2023-01-11

Geography: Canada, Province or territory, Census division
https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710013901

Census Subdivision: https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710014201


#### read data


```{r data_read}
# all region types
data1 <- read_csv(here::here("data_source", "Comox_Valley_Population_Projections_2024-02-04.csv"))

```

## Comox Valley data

```{r}

# total population
df_rd_pop_total <- data1 |> 
  filter(Gender == "T") |> 
  select(Region:Total)

df_rd_pop_total <- df_rd_pop_total |> 
  group_by(Region) |> 
  mutate(pop_change = Total - lag(Total)) |> 
  mutate(pop_change_rate = pop_change / lag(Total) *100)

df_rd_pop_total


#file_name <- glue::glue("pop_northbc_rd_{today()}.rds")
#write_rds(df_rd_pop_total, here::here("data_output", file_name))

#df_rd_pop_total <- read_rds(here::here("data_output", file_name))


```

How fast is the Comox Valley RD growing?

```{r}
# net change
df_rd_pop_total |> 
  ggplot(aes(x = Year, y = pop_change)) + 
    geom_line()


```


## 2024 over 2023

```{r}

df_rd_pop_total_current <- df_rd_pop_total |> 
  filter(Year == 2024)

df_rd_pop_total_current

```


## maps

See "population_map_northerrn.Rmd"


## population pyramid

data

```{r}

df_rd_pop_pyr <- data1 |> 
  filter(Gender %in% c("F", "M")) |> 
  select(
    `Regional District`,
    Year,
    Gender,
    `0`:`90+`
    )

df_rd_pop_pyr <- df_rd_pop_pyr |> 
  pivot_longer(-c(`Regional District`:Gender),
               names_to = "age_grp", values_to = "pop_total")

df_rd_pop_pyr

# set male values to negative
df_rd_pop_pyr <- df_rd_pop_pyr |> 
  mutate(
    pop_total = case_when(
      Gender == "M" ~ pop_total * -1,
      Gender == "F" ~ pop_total,
      TRUE ~ NA
    )
  )

```


plot

```{r}


#community_total_population <- sum(abs(df_pl_vi_pyramid_total$pop_total))
#community_total_population <- format(community_total_population, big.mark = ",", trim = TRUE)


# and plot
pl_pyramid <-  
  df_rd_pop_pyr |> 
  filter(Year == "2001") |> 
  ggplot(
    aes(x = pop_total,
           y = age_grp,
           fill = Gender
           )) + 
  geom_col() +
  theme_classic() +
  scale_fill_manual(
    values=c("#0C3343", "#EF5A27"),          # dark blue and action orange
    labels=c("men", "women")
    )  +
  xlab("population") +
  ylab(NULL)

#pl_pyramid

# fix scale numbering

pop_range_breaks <- seq(-70000, 70000, by = 10000)

pl_pyramid_total <- pl_pyramid +
  scale_x_continuous(
    breaks  = pop_range_breaks,
    #                      limits = c(min(pop_range_breaks), max(pop_range_breaks)),
    limits = c(-70000, 70000),
#   labels = abs(pop_range_breaks)
#    labels = scales::comma
   labels = abs_comma  # custom function that combines absolute value AND commas
  )  +
# add vertical white "gridlines"
  geom_vline(xintercept = pop_range_breaks, col="white") +
  labs(title = "Vancouver Island",
       subtitle = glue::glue("total population = {community_total_population}"),

       caption = "Census 2021: Statistics Canada") +
  theme(legend.position = "bottom")

pl_pyramid_total


#ggsave(here::here("plot_output/pyramid_van_isl_2021_population.jpg"),
#       width = 13.333,
#       height = 7.5
#  )



```

