---
title: "Comox Valley RD"
subtitle: "population estimates & projections: map & pyramid"
author: "Martin Monkman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# packages
library(tidyverse)
library(here)


source("mnp_defaults_ggplot.R")

# absolute value AND comma function
# from https://stackoverflow.com/questions/37949473/how-do-you-make-ggplot-labels-use-both-labels-abs-and-labels-comma
abs_comma <- function (x, ...) {
  format(abs(x), ..., big.mark = ",", scientific = FALSE, trim = TRUE)
}




```

## read data

Source: BC Stats, PEOPLE 2020

* before the pandemic threw the immigration dimension of migration out of whack

#### read data


```{r data_read}
# all region types
#data1 <- read_csv(here::here("data_source", "Comox_Valley_Population_Projections_5yr_2024-02-04.csv"))

data1 <- read_csv(here::here("data_source", "PEOPLE 2020", "CHSA1f-Revised-Dec2020.csv"))

data1

```

## Comox Valley data

```{r}

# total population
df_reg_pop_total <- data1 |> 
  filter(str_detect(ID, "^431"))

# annual summary table
df_reg_pop_total_summary <- df_reg_pop_total |> 
  filter(Gender == "Total") |> 
  select(Region, Year, Total, `Mean age`, `Median age`) |> 
  group_by(Year) |> 
  summarise(total_pop = sum(Total))

df_reg_pop_total_summary <- df_reg_pop_total_summary |> 
  mutate(pop_change = total_pop - lag(total_pop)) |> 
  mutate(pop_change_rate = pop_change / lag(total_pop) *100)

df_reg_pop_total_summary


#file_name <- glue::glue("pop_northbc_rd_{today()}.rds")
#write_rds(df_reg_pop_total, here::here("data_output", file_name))

#df_reg_pop_total <- read_rds(here::here("data_output", file_name))


```

How fast is the Comox Valley RD growing?

```{r}
# net change
df_reg_pop_total_summary |> 
  ggplot(aes(x = Year, y = pop_change)) + 
    geom_line()


# total population

ch_total_pop <- df_reg_pop_total_summary |> 
  ggplot(aes(x = Year, y = total_pop)) + 
    geom_line(size = 2, colour = "#002060")

ch_total_pop

```

```{r}
# make it pretty

ch_total_pop +
  mnp_chart_theme +
  scale_y_continuous(
    limits = c(0, 125000),
    breaks = c(0, 25000, 50000, 75000, 100000, 125000),
    labels = scales::comma
  ) +
  labs(
    title = "Comox Valley Regional District",
    subtitle = "Population growth, 2001-2046",
    caption = "Source: BCStats & Statistics Canada"
  ) +
  xlab(NULL) +
  ylab("Total population")



#ggsave(here::here("plot_output/pyramid_van_isl_2021_population.jpg"),
#       width = 13.333,
#       height = 7.5
#  )


#ggsave(here::here("plot_output","comox_valley_population_2001-46.jpg"))




```



## 2024 over 2023

```{r}

df_reg_pop_total_current <- df_reg_pop_total |> 
  filter(Year == 2024)

df_reg_pop_total_current

```


## maps

See "population_map_northerrn.Rmd"


## population pyramid

data

```{r}
df_rd_pop_main <- data1 |> 
  filter(str_detect(ID, "^431")) 

df_rd_pop_main <- df_rd_pop_main |> 
# merge LT1 + 1-4 into 0-4
#  mutate(`0 to 4` = LT1 + `1 to 4`) |> 
  select(
    Region,
    Year,
    Gender,
    `Age 0`:`Sum 90+`
    )

df_rd_pop_main

```


```{r}
df_rd_pop_main <- df_rd_pop_main |> 
  pivot_longer(-c(Region:Gender),
               names_to = "age_grp", values_to = "pop_total")  |> 
  mutate(age_grp = str_remove(age_grp, "Age ")) |> 
  mutate(age_grp = str_remove(age_grp, "Sum ")) |> 
  mutate(age_grp = as_factor(age_grp))

levels(df_rd_pop_main$age_grp)
```

set up for pop pyramid

```{r}
df_rd_pop_pyr <- df_rd_pop_main |> 
  filter(Gender %in% c("Female", "Male")) |> 
# set male values to negative
  mutate(
    pop_total = case_when(
      Gender == "Male" ~ pop_total * -1,
      Gender == "Female" ~ pop_total,
      TRUE ~ NA
    )
  )

df_rd_pop_pyr

#write_csv(df_rd_pop_pyr, here::here("data_output", "df_rd_comox_pop_pyr.csv"))

```




### over 65

```{r}

df_rd_over65 <- data1 |> 
  filter(str_detect(ID, "^431")) |>  
  filter(Gender == "Total") |>  
  select(Region, Year, Total, `Age 65`:`Sum 90+`) 


df_rd_over65 <- df_rd_over65 |>
  group_by(Region, Year) |>
  rowwise() |> 
  mutate(`65+` = 
           sum(c_across(`Age 65`:`Sum 90+`)))


df_rd_over65_summary <- df_rd_over65 |> 
  group_by(Year) |> 
  summarise(
    total_pop = sum(Total),
    total_65plus = sum(`65+`)
  ) |> 
  mutate(pct_65plus = total_65plus / total_pop)

df_rd_over65_summary

```

percent over 80

```{r}

df_rd_over80 <- data1 |> 
  filter(str_detect(ID, "^431")) |>  
  filter(Gender == "Total") |>  
  select(Region, Year, Total, `Age 80`:`Sum 90+`) 


df_rd_over80 <- df_rd_over80 |>
  group_by(Region, Year) |>
  rowwise() |> 
  mutate(`80+` = 
           sum(c_across(`Age 80`:`Sum 90+`)))


df_rd_over80_summary <- df_rd_over80 |> 
  group_by(Year) |> 
  summarise(
    total_pop = sum(Total),
    total_80plus = sum(`80+`)
  ) |> 
  mutate(pct_80plus = total_80plus / total_pop)

df_rd_over80_summary

```

join 65+ and 80+ tables

```{r}

df_rd_over6580 <- df_rd_over65_summary |> 
  left_join(df_rd_over80_summary, by = c("Year", "total_pop"))

df_rd_over6580_total <- df_rd_over6580 |> 
  select(Year, total_65plus, total_80plus) |> 
  pivot_longer(-Year, names_to = "age_grp", values_to = "total_of_pop")

df_rd_over6580_pct <- df_rd_over6580 |> 
  select(Year, pct_65plus, pct_80plus) |> 
  pivot_longer(-Year, names_to = "age_grp", values_to = "pct_of_pop")

df_rd_over6580_pct

```

### 65+ pct chart

```{r}
ch_over6580 <- ggplot(
  df_rd_over6580_pct,
  aes(x = Year, y = pct_of_pop, colour = age_grp)) +
  geom_line(size = 2)

ch_over6580

```

```{r}

ch_over6580 +
  mnp_chart_theme +
  labs(title = "Comox Valley: percent of population 65+",
#       subtitle = glue::glue("total population = {community_total_population}"),

       caption = "BC Stats & Statistics Canada") +
  theme(legend.position = "bottom") +
  scale_fill_manual(
    values=c("#0C3343", "#EF5A27"),          # dark blue and action orange
    labels=c("65+", "80+")
    )  +
  xlab(NULL) +
  ylab("percent of population") +
  scale_y_continuous(
    limits = c(0, 0.4),
#    breaks = c(0, 25000, 50000, 75000, 100000, 125000),
    labels = scales::label_percent()
  ) 
  
  
#ggsave(here::here("plot_output", "comox_valley_pct_over65.jpg"))

```

### 65+ total chart

```{r}
ch_over6580 <- ggplot(
  df_rd_over6580_total,
  aes(x = Year, y = total_of_pop, colour = age_grp)) +
  geom_line(size = 2)

ch_over6580

```

```{r}

ch_over6580 +
  mnp_chart_theme +
  labs(title = "Comox Valley: population 65+ and 80+",
#       subtitle = glue::glue("total population = {community_total_population}"),

       caption = "BC Stats & Statistics Canada") +
  theme(legend.position = "bottom") +
  scale_colour_manual(
    name = "age group",
    values = c("#0C3343", "#EF5A27"),          # dark blue and action orange
    labels = c("65+", "80+")
    )  +
  xlab(NULL) +
  ylab("population") +
  scale_y_continuous(
    limits = c(0, 30000),
    breaks = c(0, 10000, 20000, 30000, 40000),
    labels = scales::label_comma()
  ) 
  
  
#ggsave(here::here("plot_output", "comox_valley_total_over65.jpg"))

```



### population pyramid



plot

```{r}
#community_total_population <- sum(abs(df_pl_vi_pyramid_total$pop_total))
#community_total_population <- format(community_total_population, big.mark = ",", trim = TRUE)

plot_year <- "2023"

# and plot
df_rd_pop_pyr_1yr <- df_rd_pop_pyr |> 
  filter(Year == plot_year)

df_rd_pop_pyr_1yr

```


```{r}
pl_pyramid <-  
  ggplot(
    df_rd_pop_pyr_1yr,
    aes(x = pop_total,
           y = age_grp,
           fill = Gender
           )) + 
  geom_col() +
  theme_classic() +
  scale_fill_manual(
    values=c("#0C3343", "#EF5A27"),          # dark blue and action orange
    labels=c("men", "women")
    )  +
  xlab("population") +
  ylab(NULL)

pl_pyramid

# fix scale numbering
```


### plot labels

everysecond <- function(x){
  x <- sort(unique(x))
  x[seq(2, length(x), 2)] <- ""
  x
}

ggplot(data = df) + 
  geom_col(mapping = aes(x = x, y = y)) + 
  scale_x_discrete(labels = everysecond(df$x))




```{r}
pop_range_breaks <- seq(-1000, 1000, by = 250)

pl_pyramid_total <- pl_pyramid +
  scale_x_continuous(
    breaks  = pop_range_breaks,
    #                      limits = c(min(pop_range_breaks), max(pop_range_breaks)),
    limits = c(-1000, 1000),
#   labels = abs(pop_range_breaks)
#    labels = scales::comma
   labels = abs_comma  # custom function that combines absolute value AND commas
  )  +
  scale_y_discrete(
    breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80)
  ) +
# add vertical white "gridlines"
  geom_vline(xintercept = pop_range_breaks, col="white") +
  labs(title = glue::glue("Comox Valley: {plot_year}"),
#       subtitle = glue::glue("total population = {community_total_population}"),

       caption = "BC Stats & Statistics Canada") +
  theme(legend.position = "bottom")

pl_pyramid_total


#ggsave(here::here("plot_output/pyramid_van_isl_2021_population.jpg"),
#       width = 13.333,
#       height = 7.5
#  )



```

