---
title: "population map & pyramid"
subtitle: "Comox Valley RD"
author: "Martin Monkman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# packages
library(tidyverse)
library(here)

# absolute value AND comma function
# from https://stackoverflow.com/questions/37949473/how-do-you-make-ggplot-labels-use-both-labels-abs-and-labels-comma
abs_comma <- function (x, ...) {
  format(abs(x), ..., big.mark = ",", scientific = FALSE, trim = TRUE)
}




```

## read data

Source:

-   https://www2.gov.bc.ca/gov/content/data/statistics/people-population-community/population/population-estimates

-   https://www2.gov.bc.ca/assets/gov/data/statistics/people-population-community/population/pop_municipial_subprov_areas.xlsx

Source: BC Stats, Population Estimates by Regional Districts and Development Regions

Population Distribution Vancouver Island and Coast by Age Group\
2011 2016 2021\
Under 15 years old 14% 13% 13% Working Age Population (Aged 15 - 64) 67% 65% 62% Population 65+ 19% 22% 25%

Source: BC Stats, Population by Age and Sex.

#### download

https://bcstats.shinyapps.io/popApp/

Select:

-   region type == "Regional District"

-   region(s) == as required

-   years == all

-   genders == M & F

-   age format == "Single Year Age Groups"

-   variables to display as columns = Age

Rename file as "Comox_Valley_Population_Projections_2024-02-04.csv"


NOTE: StatCan table with same data for Canada (but no projections) is 

Census Division

Population estimates, July 1, by census division, 2016 boundaries1, 2, 3, 4, 5, 6
Frequency: Annual

Table: 17-10-0139-01

Release date: 2023-01-11

Geography: Canada, Province or territory, Census division
https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710013901

Census Subdivision: https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710014201


#### read data


```{r data_read}
# all region types
#data1 <- read_csv(here::here("data_source", "Comox_Valley_Population_Projections_5yr_2024-02-04.csv"))

data1 <- read_rds(here::here("data_output", "pop_bcdata_devreg_2024-02-04.rds"))

```

## Vancouver Island data

```{r}

# total population
df_vanisle_pop_total <- data1 |> 
  filter(Region.Name == "Vancouver Island/Coast") |> 
  filter(Gender == "T") |> 
  select(Region:Total)

df_vanisle_pop_total <- df_vanisle_pop_total |> 
  group_by(Region) |> 
  mutate(pop_change = Total - lag(Total)) |> 
  mutate(pop_change_rate = pop_change / lag(Total) *100)

df_vanisle_pop_total


#file_name <- glue::glue("pop_northbc_rd_{today()}.rds")
#write_rds(df_devreg_pop_total, here::here("data_output", file_name))

#df_devreg_pop_total <- read_rds(here::here("data_output", file_name))


```

How fast is Vancouver Island growing?

```{r}
# net change
df_vanisle_pop_total |> 
  ggplot(aes(x = Year, y = pop_change)) + 
    geom_line()


```


## data set up

long-form data for population pyramid

* M is set to negative for plotting

* age group is a factor for correct plot order

```{r}

## population pyramid

# data

df_pop_pyr <- data1 |> 
  filter(Gender %in% c("F", "M")) 

df_pop_pyr <- df_pop_pyr |> 
# merge LT1 + 1-4 into 0-4
#  mutate(`0 to 4` = LT1 + `1 to 4`) |> 
  select(
    Region.Name,
    Year,
    Gender,
    `0`:`90+`
    )  

df_pop_pyr <- df_pop_pyr |> 
  pivot_longer(-c(Region.Name:Gender),
               names_to = "age_grp", values_to = "pop_total")  |> 
  mutate(age_grp = as_factor(age_grp))

levels(df_pop_pyr$age_grp)

# set male values to negative
df_pop_pyr <- df_pop_pyr |> 
  mutate(
    pop_total = case_when(
      Gender == "M" ~ pop_total * -1,
      Gender == "F" ~ pop_total,
      TRUE ~ NA
    )
  )

df_pop_pyr

```


plot single year / region

```{r}
# set year and region
plot_year <- 2044

#plot_region <- "British Columbia"
plot_region <- "Vancouver Island/Coast"


# and plot
df_pop_pyr_1yr <- df_pop_pyr |> 
  filter(Region.Name == plot_region) |> 
  filter(Year == plot_year)

df_pop_pyr_1yr

```


```{r}
pl_pyramid <-  
  ggplot(
    df_pop_pyr_1yr,
    aes(x = pop_total,
           y = age_grp,
           fill = Gender
           )) + 
  geom_col() +
  theme_classic() +
  scale_fill_manual(
    values=c("#0C3343", "#EF5A27"),          # dark blue and action orange
    labels=c("men", "women")
    )  +
  xlab("population") +
  ylab(NULL)

pl_pyramid

# fix scale numbering
```


### plot labels



```{r}
#pop_range_breaks <- seq(-1000, 1000, by = 250)

pl_pyramid_total <- pl_pyramid +
  scale_x_continuous(
#    breaks  = pop_range_breaks,
    #                      limits = c(min(pop_range_breaks), max(pop_range_breaks)),
#    limits = c(-1000, 1000),
#   labels = abs(pop_range_breaks)
#    labels = scales::comma
   labels = abs_comma  # custom function that combines absolute value AND commas
  )  +
  scale_y_discrete(
    breaks = c('0', '10', '20', '30', '40', '50', '60', '70', '80', '90+')
  ) +
# add vertical white "gridlines"
#  geom_vline(xintercept = pop_range_breaks, col="white") +
  labs(
    title = glue::glue("{plot_region}: {plot_year}"),
#       subtitle = glue::glue("total population = {community_total_population}"),

       caption = "BC Stats & Statistics Canada"
       ) +
  theme(legend.position = "bottom")

pl_pyramid_total


ggsave(here::here(glue::glue("plot_output/pyramid_{plot_region}_{plot_year}_population.jpg")))

```

