---
title: "population map"
author: "Martin Monkman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# packages
library(tidyverse)
library(here)

# create list of Northern BC Census Divisions (== Regional Districts) numbers

# Cariboo [41], Fraser Fort George [53], 
# Kitimat-Stikine [49], Skeena-Queen Charlotte [North Coast 47], 
# Bulkley-Nechako [51], Stikine [57], 
# Northern Rockies [59], Peace River [55]
# and parts of Squamish-Lillooet and North Okanagan

cd_list <- c("41", "53", "49", "47", "51", "57", "59", "55")
cd_list <- as.numeric(cd_list) * 1000

```

## read data

Source:

-   https://www2.gov.bc.ca/gov/content/data/statistics/people-population-community/population/population-estimates

-   https://www2.gov.bc.ca/assets/gov/data/statistics/people-population-community/population/pop_municipial_subprov_areas.xlsx

Source: BC Stats, Population Estimates by Regional Districts and Development Regions

Population Distribution Vancouver Island and Coast by Age Group\
2011 2016 2021\
Under 15 years old 14% 13% 13% Working Age Population (Aged 15 - 64) 67% 65% 62% Population 65+ 19% 22% 25%

Source: BC Stats, Population by Age and Sex.

#### download

https://bcstats.shinyapps.io/popApp/

Select:

-   region type == "Regional District"

-   region(s) == as above

-   years == all

-   genders == Totals

-   age format == "5-year Age Groups"

-   variables to display as columns = Age

Rename file as "pop_regional district_est proj_5 yr.csv"


NOTE: StatCan table with same data for Canada (but no projections) is 

Census Division

Population estimates, July 1, by census division, 2016 boundaries1, 2, 3, 4, 5, 6
Frequency: Annual

Table: 17-10-0139-01

Release date: 2023-01-11

Geography: Canada, Province or territory, Census division
https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710013901

Census Subdivision: https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710014201


#### direct pull!

Uses BC Data Catalogue


```{r data_pull, eval=FALSE}

resource_codes <-  bcdata::bcdc_tidy_resources("86839277-986a-4a29-9f70-fa9b1166f6cb") %>% pull(id)

data1 <- map_dfr(
resource_codes,
bcdata::bcdc_get_data,
record = "86839277-986a-4a29-9f70-fa9b1166f6cb",
show_col_types = FALSE
)

file_name <- glue::glue("pop_bcdata_{today()}.rds")
write_rds(data1, here::here("data_output", file_name))

```

```{r data_read}
file_name <- glue::glue("pop_bcdata_{today()}.rds")

data1 <- read_rds(here::here("data_output", file_name))

```

explore the data 

```{r}

df_rd_pop_total <- data1 |> 
  filter(Region.Type == "Regional District") |> 
  filter(Region %in% cd_list) |> 
  filter(Gender == "T") |> 
  select(Region:Total)

# transform regional district number to conform to mapping data
df_rd_pop_total <- df_rd_pop_total |> 
  mutate(geo_uid = glue::glue("59{Region}")) |> 
  mutate(geo_uid = str_sub(geo_uid, 1, 4)) |> 
  relocate(geo_uid)

df_rd_pop_total <- df_rd_pop_total |> 
  group_by(Region) |> 
  mutate(pop_change = Total - lag(Total)) |> 
  mutate(pop_change_rate = pop_change / lag(Total) *100)

df_rd_pop_total


file_name <- glue::glue("pop_northbc_rd_{today()}.rds")
write_rds(df_rd_pop_total, here::here("data_output", file_name))

df_rd_pop_total <- read_rds(here::here("data_output", file_name))


```

How fast is the Cariboo RD growing?

```{r}

df_rd_pop_total |> 
  filter(Region.Name == "Cariboo") |> 
  ggplot(aes(x = Year, y = pop_change)) + 
    geom_line()


```


## 2022 over 2021

```{r}

df_rd_pop_total_current <- df_rd_pop_total |> 
  filter(Year == 2022)

df_rd_pop_total_current

```


## link shapes


This first section is plagiarized from another project--I wanted to make sure I had a working reference.

```{r}
# retrieve sf dataframe

#cd_pop_shape <- get_census(dataset='CA21', regions=list(PR="59"),
##                         vectors=c("median_at_cfc"="v_CA16_2456"), #omit
#                         level='CD', quiet = TRUE, 
#                         geo_format = 'sf', labels = 'short')
#
#cd_pop_shape <- cd_pop_shape %>% 
#  janitor::clean_names() %>%
#  # calculate household size
#  mutate(household_size = population/households)

#write_rds(cd_pop_shape, here("data_geography", "sf_2021_cd_bc_pop.rds"))


# load dataframe created in the now-commented-out code above
cd_pop_shape <- read_rds(here::here("data_geography", "sf_2021_cd_bc_pop.rds"))
```

basic map of household size

```{r}

ggplot(cd_pop_shape) + 
  geom_sf(aes(fill = household_size))

```


```{r}
# NOTE: starting with the dataframe that has the data but not the shape does not include the "geometry" variable we need for mapping! 


df_pop_shape <- cd_pop_shape %>% 
  left_join(df_rd_pop_total_current, by = "geo_uid") 

# drop names of RDs that our out-of-scope
df_pop_shape <- df_pop_shape |> 
    filter() |> 

  mutate(name_rd = case_when(
    Region %in% cd_list ~ name,
    TRUE ~ NA_character_
  )) |> 
  # clean up names
  mutate(name_rd = str_remove(name_rd, " \\(RD\\)$")) |> 
  mutate(name_rd = str_remove(name_rd, " \\(REG\\)$"))



# create sf object with only in-scope RDs
df_sf_rd_list <- df_pop_shape |> 
    filter(!is.na(name_rd))

df_sf_rd_list

```

## map growth rate




```{r map_vet_orange}

map_rd_pop_growth <- ggplot(df_pop_shape) +
  geom_sf(aes(fill = pop_change_rate), colour = "grey90") +
  scale_fill_gradient(
    low = "#eef0f2",
    high = "#0c3343",
    na.value = "#ffffff",
    name = "Percent change"
    ) +
  coord_sf(
  # projection
    datum = sf::st_crs(3153)
  # crop
  # longitude
#    xlim = c(-123.4,-128.8),
  # latitude
#    ylim = c(48.4, 50.8)
    ) +  
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = c(0.1, 0.4)
  ) +
#  coord_sf(datum = sf::st_crs(3153))
#  coord_sf(datum = NULL)
  labs(title = "Population growth, 2021–2022", 
       subtitle = "Northern BC Regional Districts",
       caption = "Source: Statistics Canada & BC Stats",
       x = NULL,
       y = NULL)

map_rd_pop_growth

map_rd_pop_growth +
  # add regional district name
  geom_sf_label(
    data = df_pop_shape,
    aes(label = name), 
    colour = "black", 
    size = 1
    )  

map_rd_pop_growth +
  # add regional district name
  geom_sf_label(
    data = df_sf_rd_list,
    aes(label = name_rd ), 
    colour = "black", 
    size = 1,
    show.legend = FALSE
    )  



#ggsave(here::here("plot_output", "vi_rd_pop_growth.jpg"))
ggsave(here::here("plot_output", "rd_pop_growth_600dpi_test.jpg"), dpi = 600)

```


#### map with just the labels


```{r}
ggplot() +
  geom_sf_text(
    data = df_sf_rd_list, 
    aes(label = name_rd ), 
    colour = "#000000", 
    size = 1.5,
    show.legend = FALSE
    )  

```


### generalized loop through a few years


```{r}
df_rd_pop_total |> 
  filter(Year == 2014) |> 
  summarise(min(pop_change_rate, na.rm = TRUE),
            max(pop_change_rate, na.rm = TRUE))
```


#### loop

```{r}

year_range <- 2015:2022

for (i in year_range) {

print(i)
getwd()
  
year_current <- i
#year_current <- 2015

year_before <- year_current - 1

chart_title <- glue::glue("Population growth, {year_before}–{year_current}")

df_rd_pop_total_current <- df_rd_pop_total |> 
  filter(Year == year_current)

#df_rd_pop_total_current


# NOTE: starting with the dataframe that has the data but not the shape does not include the "geometry" variable we need for mapping! 

df_pop_shape <- cd_pop_shape %>% 
  left_join(df_rd_pop_total_current, by = "geo_uid") 

ggplot(df_pop_shape) +
  geom_sf(aes(fill = pop_change_rate), colour = "grey95") +


# LABELS    
#  geom_sf_text(
  geom_sf_label(
    data = df_sf_rd_list, 
    aes(label = name_rd), 
    colour = "#000000", 
    size = 1.5,
    show.legend = FALSE
    )  +


#  scale_fill_gradient(
#      low = "#a6611a",
#    high = "#018571",
#    na.value = "#ffffff",
#    limits = c(-2.0, 3.0),
#    name = "Percent change"
  scale_fill_distiller(
    name = "Percent change\nfrom previous year",
    palette = "BrBG", 
    direction = 1,
    na.value = "grey90",
    limits = c(-2, 3)
    ) +
  coord_sf(
  # crop
  # longitude
#    xlim = c(-123.4,-128.8),
  # latitude
#    ylim = c(48.4, 50.8),
  # projection
#    datum = sf::st_crs(3005)
    datum = sf::st_crs(3153)
    ) +  
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = c(0.1, 0.4)
  ) +
#  coord_sf(datum = sf::st_crs(3153))
#  coord_sf(datum = NULL)

  labs(title = chart_title, 
       subtitle = "Northern BC Regional Districts",
       caption = "Source: Statistics Canada",
       x = NULL,   # need to NULL so that there are no "X" and "Y" artifacts
       y = NULL
       )
       

#ggsave(here::here("plot_output", "vi_rd_pop_growth.jpg"))
file_name <- glue::glue("rd_pop_growth_{year_current}.jpg")
ggsave(here::here("plot_output", file_name), 
       dpi = 600,
       width = 9,
       height = 6,
       units = "in"
       )

}

```


```{r}

year_range <- 2015:2022

for (i in year_range) {
  print(i)
}

```





#### population projections

```{r}

year_range <- 2023:2024

for (i in year_range) {

print(i)
getwd()
  
year_current <- i
year_before <- year_current - 1

chart_title <- glue::glue("Population growth projection, {year_before}–{year_current}")

df_rd_pop_total_current <- df_rd_pop_total |> 
  filter(Year == year_current)

#df_rd_pop_total_current


# NOTE: starting with the dataframe that has the data but not the shape does not include the "geometry" variable we need for mapping! 

df_pop_shape <- cd_pop_shape %>% 
  left_join(df_rd_pop_total_current, by = "geo_uid") 

ggplot(df_pop_shape) +
  geom_sf(aes(fill = pop_change_rate), colour = "grey90") +
#  scale_fill_gradient(
#      low = "#a6611a",
#    high = "#018571",
#    na.value = "#ffffff",
#    limits = c(-2.0, 3.0),
#    name = "Percent change"
  scale_fill_distiller(
    name = "Percent change\nfrom previous year",
    palette = "BrBG", 
    direction = 1,
    na.value = "grey90",
    limits = c(-2, 3)
    ) +
  coord_sf(
  # projection
    datum = sf::st_crs(3153),
  # crop
  # longitude
    xlim = c(-123.4,-128.8),
  # latitude
    ylim = c(48.4, 50.8)
    ) +  
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = c(0.1, 0.4)
  ) +
#  coord_sf(datum = sf::st_crs(3153))
#  coord_sf(datum = NULL)

  labs(title = chart_title, 
       subtitle = "Vancouver Island Regional Districts",
       caption = "Source: BC Stats")

#ggsave(here::here("plot_output", "vi_rd_pop_growth.jpg"))
file_name <- glue::glue("vi_rd_pop_growth_{year_current}.jpg")
ggsave(here::here("plot_output", file_name), 
       dpi = 600,
       width = 9,
       height = 6,
       units = "in"
       )

}

```
